---
title: "West Coast EEZs"
author: "Sofia Sarak"
format: pdf
editor: visual
---

```{r}
#| message: false
library(stars)
library(terra)
library(sf)
library(here)
library(tmap)
library(tidyverse)
```

## Loading in Data

SST

```{r}
# There are 5 raster files of annual sst, each for a different year
# Their naming convention is the same, so we can use a for loop to load them all in at once

# Create a vector of the years we have rasters for (what we will iterate over)
years <- c(2008, 2009, 2010, 2011, 2012)

# Initialize empty vectors for for loop
raster_names <- c()
file_paths <- c()

# for loop follows the following steps:
#   - iterates over the sequence in years vector
#   - for each year, creates a variable name including the year
#   - for each year, creates a file path
#   - assigns the variable name to the raster, which is read in via the created file path
for (i in seq(years)){                                      
  raster_names[i] <- paste0("sst_", years[i])
  file_paths[i] <- paste0("data/average_annual_sst_", years[i], ".tif")
  
  assign(raster_names[i], rast(here(file_paths[i])))
}

rm(raster_names, file_paths, i, years)
```

providing elevation data, in meters

Bathymetry

```{r}
# Use terra package to load in raster
depth <- rast(here("data/depth.tif"))
```

Exclusive Economic Zones

```{r}
# Use stars package to load in shape data frame
# quiet = TRUE argument removes output/message
eez <- st_read(here("data/wc_regions_clean.shp"), quiet = TRUE)
```

## Preparing Data

In order to proper perform analyses using all of data sets, we first need to ensure that they all have matching CRSs.

First, we are going to stack our five SST rasters into one. As they all come from the same source and only differ in year we can reasonably assume that they all have the same CRS, but stacking would also throw an error if that was not the case.

```{r}
# Create raster stack
sst <- c(sst_2008, sst_2009, sst_2010, sst_2011, sst_2012)
```

We can now compare our 3 data frames to see if they all have the same CRSs.

```{r}
# sst and depth rasters
if (crs(sst)==crs(depth)){
  print("CRSs of sst and depth match!")
}else{
  print("CRSs of sst and depth do not match, transforming to CRS of sst")
  depth <- project(depth, crs(sst))
}

# sst raster and eez object
if (crs(sst)==crs(eez)){
  print("CRSs of sst and eez match!")
}else{
  print("CRSs of sst and eez do not match, transforming to CRS of sst")
  eez <- st_transform(eez, crs(sst))
}

# We transfromed depth and eez to match the CRS of sst -- did it work?
if (crs(depth)==crs(eez)){
  print("CRSs of sst and eez match (transformations worked!)")
}else{
  print("Something went wrong...")
}
```

## Processing Data

We want to combine our sst and depth data, but in order to do so, we have to ensure they have the same extent, resolution, and position.

There is no method to extract position, but by matching up extent and resolution we often fix position as well (CONFIRM).

```{r}
# Is the extent the same?
ext(sst) == ext(depth)

# Is the resolution the same?
res(sst) == res(depth)
```

Find mean SST over the years 2008-2012.

```{r}
# mean() operation on raster of multiple layers creates object with one raster layer
mean_sst <- mean(sst)
```

From the metadata (CONFIRM), we know that SST is stored in Kelvin instead of Celsius. In order to make our results more comprehensible, we will convert our data to Celsius.

```{r}
# Subtract 273.15 (conversion from K to C) from all pixels
mean_sst_c <- mean_sst - 273.15
```

In order to ensure that `depth` and `sst` have the same extent, we will crop `depth` to match `sst`.

```{r}
# Crop depth raster to limits of sst raster
depth_cropped <- crop(depth, y = mean_sst_c)
```

To fix the issue of mismatched resolutions, we will resample the depth data to match the resolution of `sst`.

Resampling is the method by which we can compute values for new pixel locations in `depth` based on custom resolutions and origins. We will use the "nearest neighbor approach", a technique in which the value of each cell in the output raster (in this case, `depth_cropped`) is calculated using the value of the nearest cell in the input raster (`mean_sst`) ([ESRI GIS Dictionary](https://support.esri.com/en-us/gis-dictionary/nearest-neighbor-resampling)). This fills in the new cells that will be created in order for `depth` to match the resolution of `sst`.

```{r}
# Resample depth data to match specifications of sst data, specifying nearest neighbor method
depth_cropped_resampled <- resample(depth_cropped, y = mean_sst_c, method = "near")
```

A way to know if we have succeeded in matching the extent, resolution, and position of our two rasters is by seeing if they will stack or not. If they do not completely match, an error will be thrown and stacking will fail.

```{r}
depth_sst <- c(depth_cropped_resampled, mean_sst_c)
```

We successfully stacked our two rasters!


## Finding areas with suitable growing conditions: oysters.

Research has shown that oysters need the following conditions for optimal growth:

sea surface temperature: 11-30°C depth: 0-70 meters below sea level (or -70 in our data)

These specifications come from the EDS223 assignment description, but can also be retrieved from resources such as [SeaLifeBase](https://www.sealifebase.ca/search.php).

In order to find the areas with suitable growing conditions for oysters, we will assign a value of 1 to each pixel that is within the suitable range and 0 otherwise, for both of our separate raster layers.

```{r}
# Change layer names to be more indicative ("mean" --> "sst_mean")
names(depth_sst) <- c("depth", "sst_mean_c")

# Create copy of data frame to serve as mask
depth_sst_mask <- depth_sst
```

```{r}
sst_suitability <- matrix(c(-Inf, 11, 0,
                          11, 30, 1, 
                          30, Inf, 0),
                          nrow = 3, byrow = TRUE)

depth_sst_mask[["sst_mean_c"]] <- classify(depth_sst_mask[["sst_mean_c"]], rcl = sst_suitability)
```

```{r}
depth_suitability <- matrix(c(-Inf, -70, 0,
                              -70, 0, 1,
                              0, Inf, 0),
                            nrow = 3, byrow = TRUE)

depth_sst_mask[["depth"]] <- classify(depth_sst_mask[["depth"]], rcl = depth_suitability)
```

```{r}

```

\*\* when writing generalizable function, arguments would be: min_sst, max_sst, min_depth, max_depth and replace the values in the matrix with the parameters in the function

To create one layer that contains suitability based on *both* variables, we will multiple the two layers, obtaining the following results:

-   if a cell = 1, it is suitable based on both conditions
-   if a cell = 0, it is not suitable (either based on one of or both conditions)

This can be true because anything multiplied by 0 (not suitable in this case), will equal 0.

-   can do via boolean:
-   if depth == 0 and sst == 0, then multiply
-   if depth == 0 and sst = 1, then multiple if depth

```{r}
oyster_suitability <- depth_sst_mask[["depth"]] * depth_sst_mask[["sst_mean_c"]]

names(oyster_suitability) <- c("suitable")
```


## Determining the most suitable EEZ

We want to determine the total suitable area within each EEZ in order to rank zones by priority. 

This requires three distinct steps:
  1. Find the number of cells within each zone that are deemed suitable (based on previous analysis).
  2. Find the physical area that each cell represents (in km^2).
  3. Multiply number of cells by area that cells represent to get actual suitable area, by zone.

```{r}
# Create a raster of species suitability only within the EEZ boundaries
suitable_westcoast <- mask(oyster_suitability, eez)

# Rasterize EEZ (will allow for zonal operations)
eez_rast <- rasterize(eez, suitable_westcoast,
                      field = "rgn")      # Set region as the value we want populating the eez raster
```

We can find the number of suitable *cells* in each EEZ region by using the `terra::zonal` function. It summarizes values of one SpatRaster based on the zones provided by another. In our case, we will be summing the values of our suitable cells grouped by eez.

Summation works to provide the number of suitable cells in this case because we had populated suitable with "1"s and unsuitable with "0"s.

```{r}
# Find the number of suitable cells per EEZ region
suitable_per_region <- zonal(suitable_westcoast, eez_rast, fun = "sum", na.rm = TRUE)
```

```{r}
#| echo: false
# Create clean table to view sums
suitable_per_region %>% 
  knitr::kable(col.names = c("EEZ Region", "# of Suitable Cells"))
```

Find area of each cell in square km.
```{r}
# cellSize() returns a SpatRaster object of information, we want to extract just area
cell_area <- cellSize(suitable_westcoast, unit="km")[1]
```

Multiply cell area by number of cells.
```{r}
# Multiplying by each region by adding and specifying columns to data frame
suitable_per_region <- suitable_per_region %>% 
  mutate(cell_area = cell_area,
         area_suitable = suitable * cell_area) %>% 
  relocate(cell_area, .after = suitable) 
```

In order to produce a map of West Coast EEZs colored by amount of suitable species area in each region, we need to join our EEZ geometries to our `suitable_per_region` data frame.

```{r}
# Join eez data frame (with goemetries) to data frame with suitable area by region
eez_suitability_geom <- left_join(eez, suitable_per_region, by = "rgn")
```

The `eez_suitability_geom` now contains the 5 West Coast region polygons with their corresponding suitable areas for oysters.

The final step of our oyster analysis is to create a map of EEZ regions colored by amount of suitable area. For this, we use our original depth raster to provide a background image (and general outlines of the California coast), overlayed with the 5 West Coast EEZs, colored by amount of suitable area.

```{r}
#| messages: false
#| warnings: false

# Turn off autoscale -- messes with font size
tmap_options(component.autoscale = FALSE) 

# Create map using tmap
tm_shape(depth) +                                      # Plot depth raster as first layer
  tm_raster(col.scale  = tm_scale(values = c("#1d3891", "#84d18d")),    
            col.legend = tm_legend(show = FALSE)) +    # Hide depth legend because not of interest
tm_shape(eez_suitability_geom) +                       # Plot EEZ polygons
  tm_polygons(fill = "area_suitable",                  # Fill by suitable area
              fill.scale = tm_scale(values = "brewer.yl_or_rd"),
              fill.legend = tm_legend(title = "Suitable Area (km^2)"),
              fill_alpha = 0.8,
              lwd = 1.5) + 
  tm_text(text = "rgn",                                # Add EEZ zone labels
          size = 0.6,
          xmod = -2.35,
          bgcol = "white",
          bgcol_alpha = 0.5,
          fontface = "bold")+
tm_grid(alpha = 0.5) +
tm_title("Suitability of West Coast EEZs for Oyster Aquaculture",
         size = 1)
```

## Finding Most Suitable EEZs for Any Species

In the code above, we developed a method of extracting the most suitable regions for marine aquaculture based on just a few parameters: temperature range and depth range. By making our workflow generalizable, we can create a function that takes a handle of arguments and produces a map of suitable West Coast EEZs for any marine species.

### Generalizing Workflow

Our function will follow a few assumptions:
- depth entered as a positive value (ex; 70 feet deep/below sea level is to be entered as 70, not -70.) 
- max_depth is the deepest


```{r}

west_coast_suitability_map <- function(species, min_sst, max_sst, min_depth, max_depth){

# Creating depth and sst mask based on min and max values
depth_sst_mask <- depth_sst

# sst matrix and classification
sst_suitability <- matrix(c(-Inf, min_sst, 0,
                          min_sst, max_sst, 1, 
                          max_sst, Inf, 0),
                          nrow = 3, byrow = TRUE)

depth_sst_mask[["sst_mean_c"]] <- classify(depth_sst_mask[["sst_mean_c"]], rcl = sst_suitability)

# depth matrix and classification
depth_suitability <- matrix(c(-Inf, -max_depth, 0,
                              -max_depth, min_depth, 1,
                               min_depth, Inf, 0),
                               nrow = 3, byrow = TRUE)

depth_sst_mask[["depth"]] <- classify(depth_sst_mask[["depth"]], rcl = depth_suitability)

# Creating suitability raster
suitability_raster <- depth_sst_mask[["depth"]] * depth_sst_mask[["sst_mean_c"]]

# Renaming raster layer value to be more accurate
names(suitability_raster) <- c("suitable")

# Create a raster of species suitability only within the EEZ boundaries
suitable_westcoast <- mask(suitability_raster, eez)

# Rasterize EEZ (will allow for zonal operations)
eez_rast <- rasterize(eez, suitable_westcoast,
                      field = "rgn")      # Set region as the value we want populating the eez raster

# Find the number of suitable cells per EEZ region
suitable_per_region <- zonal(suitable_westcoast, eez_rast, fun = "sum", na.rm = TRUE)

# cellSize() returns a SpatRaster object of information, we want to extract just area
cell_area <- cellSize(suitable_westcoast, unit="km")[1]

# Multiplying by each region by adding and specifying columns to data frame
suitable_per_region <- suitable_per_region %>% 
  mutate(cell_area = cell_area,
         area_suitable = suitable * cell_area) %>% 
  relocate(cell_area, .after = suitable) 

# Join eez data frame (with goemetries) to data frame with suitable area by region
eez_suitability_geom <- left_join(eez, suitable_per_region, by = "rgn")

# Turn off autoscale -- messes with font size
tmap_options(component.autoscale = FALSE) 

# Create map using tmap
tm_shape(depth) +                                      # Plot depth raster as first layer
  tm_raster(col.scale  = tm_scale(values = c("#1d3891", "#84d18d")),    
            col.legend = tm_legend(show = FALSE)) +    # Hide depth legend because not of interest
tm_shape(eez_suitability_geom) +                       # Plot EEZ polygons
  tm_polygons(fill = "area_suitable",                  # Fill by suitable area
              fill.scale = tm_scale(values = "brewer.yl_or_rd"),
              fill.legend = tm_legend(title = "Suitable Area (km^2)"),
              fill_alpha = 0.8,
              lwd = 1.5) + 
  tm_text(text = "rgn",                                # Add EEZ zone labels
          size = 0.6,
          xmod = -2.35,
          bgcol = "white",
          bgcol_alpha = 0.5,
          fontface = "bold")+
tm_grid(alpha = 0.5) +
tm_title(paste("Suitability of West Coast EEZs for", species, "Aquaculture"),
         size = 1)
}
```


## Running Function on Species of Choice: Bull Kelp

Talk a little about Bull kelp

```{r}
west_coast_suitability_map(species = "Bull Kelp" ,
                           min_sst = 10,
                           max_sst = 16,
                           min_depth = 40,
                           max_depth = 17)
```

